---
output: 
  pdf_document:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: ~/Dropbox/Uni/Misc/template.tex
title: "ChangeMyView"
subtitle: "Can Moral Appeals Facilitate Persuasion and Agreement?"
thanks: "**Current draft version**: `r format(Sys.time(), '%B %d, %Y')` --- PLEASE DO NOT CITE OR REDISTRIBUTE WITHOUT PERMISSION. "
#; **Corresponding author**: patrick.kraft@stonybrook.edu.; Replication files are available on the author's Github account (http://github.com/pwkraft)."
author:
- name: Patrick W. Kraft
  affiliation: Stony Brook University
abstract: "The American electorate is becoming increasingly polarized. According to research in moral psychology, these growing disagreements between liberals and conservatives can be attributed to fundamental differences in the moral frameworks that shape individual ideology. Indeed, scholars suggest that ideologues would be more likely to come to agreements if both sides spoke the same ``moral language.'' While this implicit assumption certainly has intuitive appeal, it remains largely untested empirically. Drawing on a unique dataset from the online discussion board _Reddit_, this paper examines how moral appeals can affect individual persuasion and the likelihood of agreement and consensus building through deliberation."
keywords: "Moral Foundations, Attitude Change, Persuasion, Political Agreement"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
spacing: double
bibliography: /home/patrick/Dropbox/Uni/Lit/Literature
biblio-style: /home/patrick/Dropbox/Uni/Lit/apsr2006
endnote: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE,fig.height=3, fig.width=6)
```

<!-- Notes:

- growing literature on role of morality in politics
- two different perspectives that developed somewhat independently: moral foundations theory and moral conviction
- both lead to different predictions regarding the role of morality to help reach compromise: mft suggests that focusing on the same moral foundation should lead to agreement among liberals and conservatives, whereas the moral conviction literature suggests that people are less willing to compromise if their attitudes are moralized.
- this paper tests both perspectives using the moral foundations dictionary and only finds support for the moral conviction argument.

-->


# Introduction

Recent years have witnessed a resurgence in partisan polarization in the United States. Politically engaged citizens hold more diverging policy views, are more ideologically extreme, and exhibit stronger negative affect towards out-partisans than in the past [@hetherington2001resurgent; @abramowitz2008polarization; @iyengar2012affect; @mason2014disrespectfully; @huddy2015expressive; @iyengar2015fear]. A growing literature in moral psychology---Moral Foundations Theory---attributes this divide (at least partially) to fundamental differences in moral frameworks that guide liberals and conservatives [@haidt2012righteous]. According to this view, liberals focus on _individualizing_ moral foundations, which unclude care/harm and fairness/cheating. Conservatives, on the other hand, also emphasize the remaining _binding_ foundations of loyalty/betrayal, authority/subversion, and sanctity/degradation [@haidt2007morality; @graham2009liberals; @haidt2012righteous]. Differential emphasis on these moral dimensions is systematically related to attitudes towards a wide variety of divisive political issues [e.g. @koleva2012tracing; @kertzer2014moral; @low2015moral], personality traits like individual social dominance orientation (SDO) and right-wing authoritarianism (RWA) [@federico2013mapping], as well as voting behavior [@franks2015using]. Overall, this body of research suggests that liberals and conservatives endorse different moral foundations and that these differences are related to political attitudes, evaluations, and behavior.

 




Political discussions among individuals holding diverging views can mitigate the polarizing effects of partisan motivated reasoning [see also @mendelberg2002deliberative; @eveland2011beyond]. In general, diverse social networks have been shown to reduce attitude extremity [@levitan2009social] and enhance the understanding of opposing viewpoints [@mutz2002cross]. Deliberation in particular increases political sophistication [@gastil1999increasing] and shapes attitude change towards a common consensus [@vinokur1978depolarization; @barabas2004deliberation].

@druckman2003framing directly investigated how discussions can moderate elite framing effects and showed that discussions in homogeneous groups had a polarizing effect, whereas discussions in heterogeneous groups had a depolarizing effect [see also @druckman2004political]. As such, deliberation in diverse groups can decrease the effect of polarizing elite communication. Similar research was conducted by @klar2014partisanship}, who manipulated group compositions in terms of the participants' party affiliations rather than comparing heterogeneous and homogeneous framing conditions. Her results showed that homogeneous group discussions resulted in more intense partisan motivated reasoning: individuals reported stronger preferences for the policies of their in-party and perceived them to be more effective compared to policies of their out-party. On the other hand, preferences were less extreme if the social setting was ideologically heterogeneous.

Morality could be viewed as such a potential moderator of deliberation effects. Whether individuals perceive issues in terms of their fundamental moral principles can be expected to influence the nature and tone of political discussions and therefore to moderate the downstream effects of deliberation in homogeneous and heterogeneous groups. Consider two scenarios where individuals with opposing attitudes discuss their views on abortion as compared to the appropriate tax rate. While both scenarios could result in fierce debates, we can presume that a group of pro-life and pro-choice advocates will have a harder time to find a common ground than a group of tax critics and proponents. One might argue that this difference is simply due to the nature of the issues themselves. However, how could we then explain potential differences in willingness to compromise between groups discussing the same issue? @skitka2005moral provided a possible answer to this question by introducing the concept of _moral conviction_} (or _moral mandates_) as unique features of attitude strength. In their view, moral conviction should be regarded as independent from other dimensions of attitude strength such as extremity, importance, or accessibility.

Moral convictions are characterized as attitudes that are perceived as ``absolutes, or universal standards of truth that others should also share'' [@skitka2010psychology, 269]. As such, they combine the following attributes [@skitka2010psychology]: they are viewed by individuals as applying to everyone (universality), they do not require an immediate underlying rationale but are rather seen as facts about the world (objectivity), they can be independent of authority and group norms (autonomy), they elicit strong emotional reactions, and they have an inherent motivational quality (motivation/justification). 

Building on this work, @ryan2014reconsidering argued that moral convictions are not restricted to issues that are traditionally perceived as ``moral'', such as abortion or same-sex marriage, but can also include other issues such as economic policies.

The degree of moral conviction may therefore vary between individuals as well as across issues. @ryan2014reconsidering further showed that the propensity to moralize -- i.e. the tendency to view an issue as a question of ``right and wrong'' -- is related to political participation, extreme political attitudes, arousal of negative emotions, and hostility. In a subsequent study, @ryan2016no suggested that moralization as a distinct characteristic of attitude intensity reorients behavior from maximizing gains to the general adherence to rules. Across multiple studies, the author showed that this tendency translates into stronger opposition to compromises about political issues and decreased support for compromising politicians. These patterns should also translate into attitudes towards -- and interactions with -- others who hold opposing views. Indeed, moral conviction has been shown to be related to stronger preferences for social distance from (and hostility towards) attitudinally dissimilar others and lower cooperativeness in groups holding heterogeneous views [@skitka2005moral].

The literature on moral conviction, however, is only beginning to elaborate on the political implications as the theories were mainly applied in psychology and did not explicitly consider the political context. As such, little is known about when and how individuals form moral convictions for specific political attitudes. For example, what is the role rhetoric in raising moral mandates and how does moralization affect subsequent attitude formation and belief updating? Another important question that is largely unexplored is how people react towards others who raise potentially conflicting moral considerations in social interactions? Overall then, all three literatures can benefit from an integration of the different perspectives, which is the focus of the next section.


``Across all four studies, liberal morality was primarily concerned with harm and fairness, whereas conservative moral concerns were distributed more evenly across all five foundations. These findings help explain why liberals and conservatives disagree on so many moral issues and often find it hard to understand how an ethical person could hold the beliefs of the other side: Liberals and conservatives base their moral values, judgments, and arguments on different configurations of the five foundations'' [@graham2009liberals, 1040].



# Description of Dataset
```{r}
library(tidyr)
library(dplyr)
library(purrr)
library(ggplot2)
library(quanteda)

## load data
load("out/cmv_data.Rdata")
```

```{r}
## load dictionary
dict <- dictionary(file="in/moral foundations dictionary.dic", format="LIWC")

## create corpus $ dfm for positive posts
countMFT <- function(x, dic){
  corpus <- corpus(x)
  txt <- dfm(corpus, dictionary = dic)
  data <- (data.frame(txt)/ntoken(corpus)) %>% 
    mutate(Harm = HarmVirtue + HarmVice,
           Fairness = FairnessVirtue + FairnessVice,
           Ingroup = IngroupVirtue + IngroupVice,
           Authority = AuthorityVirtue + AuthorityVice,
           Purity = PurityVirtue + PurityVice,
           MFTcombined = Harm + Fairness + Ingroup + Authority + Purity) %>%
    select(-HarmVirtue:-PurityVice)
  data
}
```

*ToDos*
=========
- clean original posts (links etc)!
- adjust confidence intervals to correct for multiple comparisons
-  add more comments in functions!


# Moral Foundations and Persuadability

```{r}
###################
### Examine OP data
###################

mft_op <- data_op[,c("title","text")] %>% 
  apply(1, paste, collapse=" ") %>%
  countMFT(dict) %>% bind_cols(data_op) %>% as_tibble()

plot_df <- mft_op %>%
  select(-name:-text, -MFTcombined) %>%
  gather(key=foundation, value=proportion,-Delta) %>%
  mutate(foundation = factor(foundation, levels = rev(c("Authority", "Ingroup", "Purity"
                                                        , "Fairness", "Harm", "MoralityGeneral"))
                             , labels = rev(c("Authority", "Ingroup", "Purity"
                                              , "Fairness", "Harm", "General\nMorality")))) %>%
  group_by(foundation,Delta) %>%
  summarise(mean = mean(proportion), sd = sd(proportion), n = n()) %>%
  mutate(se = sd/sqrt(n), cilo = mean - se, cihi = mean + se)

ggplot(plot_df, aes(y=foundation, x=mean, xmin=cilo, xmax=cihi, col=Delta, shape=Delta)) + 
  theme_classic() + theme(panel.border = element_rect(fill=NA)) + 
  geom_point(position = position_nudge(y=-.1+.2*plot_df$Delta)) + 
  geom_errorbarh(height=0, position = position_nudge(y=-.1+.2*plot_df$Delta)) + 
  ylab("Moral Foundation") + xlab("Proportion of Dictionary Terms") +
  ggtitle("Moral Foundations and Persuadability")
```


# Is Moral Consistency Convincing?

```{r}
### Examine pair data

mft_pair <- countMFT(data_pair$pos_text, dict) - countMFT(data_pair$neg_text, dict)

checkTerm <- function(x, return_data = FALSE){
  cases <- c(map(x, grep, data_pair$op_text), map(x, grep, data_pair$op_title)) %>% 
               unlist() %>% unique() %>% sort()
  
  plot_df <- mft_pair[cases,] %>% 
    gather() %>%
    mutate(key = factor(key, levels = rev(c("Authority", "Ingroup", "Purity", "Fairness", "Harm"
                                            , "MFTcombined", "MoralityGeneral"))
                        , labels = rev(c("Authority", "Ingroup", "Purity", "Fairness", "Harm"
                                         , "Foundations\nCombined", "General\nMorality")))) %>%
    group_by(key) %>%
    summarise(mean = mean(value), sd = sd(value), n = n()) %>%
    mutate(se = sd/sqrt(n), cilo = mean - se, cihi = mean + se)
  
  if(return_data) return(plot_df)
  
  xlim <- max(abs(c(plot_df$cilo,plot_df$cihi)))
  xlim <- c(-xlim,xlim)
  
  ggplot(plot_df, aes(y=key, x=mean, xmin=cilo, xmax=cihi)) + 
    theme_classic() + theme(panel.border = element_rect(fill=NA)) +
    geom_point() + geom_errorbarh(height=0) + geom_vline(xintercept = 0) +
    labs(title = paste0("Search Query (",length(cases)," hits):"), subtitle = paste(x, collapse="; ")
         , x = NULL, y = "Moral Foundation") +
    scale_x_continuous(limits=xlim, breaks=.9*xlim, labels = c("Less Persuasive", "More Persuasive")) +
    theme(axis.ticks.x=element_blank())
}

mftTerms <- function(x){
  map(x, grep, names(dict)) %>% 
    unlist() %>% unique %>% sort() %>%
    map(~gsub("*","", dict[[.]], fixed=T)) %>%
    unlist()
}
```

_Authority_

```{r}
checkTerm(mftTerms("AuthorityVirtue"))
```

```{r}
checkTerm(mftTerms("AuthorityVice"))
```

_Ingroup_

```{r}
checkTerm(mftTerms("IngroupVirtue"))
```

```{r}
checkTerm(mftTerms("IngroupVice"))
```

_Purity_

```{r}
checkTerm(mftTerms("PurityVirtue"))
```

```{r}
checkTerm(mftTerms("PurityVice"))
```

_Harm_

```{r}
checkTerm(mftTerms("HarmVirtue"))
```

```{r}
checkTerm(mftTerms("HarmVice"))
```

_Fairness_

```{r}
checkTerm(mftTerms("FairnessVirtue"))
```

```{r}
checkTerm(mftTerms("FairnessVice"))
```


# How about other topics?

```{r}
checkTerm("")
```

```{r}
checkTerm("politic")
```

```{r}
checkTerm(c("liberal","conservati"))
```

```{r}
checkTerm(c("democrat","republican"))
```

```{r}
checkTerm("climate")
```

```{r}
checkTerm(c("climate change","global warming"))
```

```{r}
checkTerm(c("politic","liberal","conservati","democrat","republican","president"
            ,"climate change","global warming"))
```

